Stable quantum memories with few measurements

### 2017-04-21 IsingECC/manuscript_ising_ecc.tex
updated latest draft

To calculate this probability, we proceed via Bayes rule: 
\begin{align}
\label{eq:bayes}
&P(d_1:d_2 | d_1^{x_1,t_1} \wedge d_2^{x_2,t_2}) = \\ \notag
&\frac{P(d_1^{x_1,t_1} \wedge d_2^{x_2,t_2} | d_1:d_2) P(d_1 : d_2)}{P(d_1^{x_1,t_1}\wedge d_2^{x_2,t_2})}
\end{align}
The individual terms on the right hand side of equation \ref{eq:bayes} are straightforward to interpret.  $P(d_1^{x_1,t_1} \wedge d_2^{x_2,t_2} | d_1:d_2)$ represents the probability that two measured defects would be at $(x_1,t_1)$ and $(x_2,t_2)$ given that they are indeed a pair.  This can be related to the probability that a one dimensional diffusion process with diffusion constant $D$ will perform an excursion of distance $|x_2-x_1|$ or greater in a time $t_2 - t_1$; \textit{i.e.,} will perform an excursion that can measurement patches at $x_1$ and $x_2$. Explicitly,
\begin{align}
\label{eq:bayesformula}
P(d_1^{x_1,t_1} \wedge d_2^{x_2,t_2} | d_1:d_2) &= \red{\int_{x_1}^{x_2} ds (.)} \nonumber \\ 
&= 1 - {\rm erf}\bigg(\frac{|x_2-x_1|}{2\sqrt{D |t_2 - t_1|}}\bigg) 
\end{align}
For our analysis, we will choose $D\propto \gamma_0$.  The exact correspondence between $D$ and $\gamma_0$ depends on the details of the error correction algorithm itself, so, in practice, we treat the constant of proportionality as an empirically tuned parameter. \red{Mention approximation that the pair was created at the center.}

@@

 The simple error model is defined as follows:  suppose that a single pair of defects is in a system, and that no new pairs will be introduced.  Further, suppose that there is an error process that occasionally increases the distance between this pair by an amount $2\sqrt{\gamma_0 \delta t}$, for hopping rate $\gamma_0$ and for system time $\delta t$.  Note that this means the error process acts on longer distance scales as system time increases. \red{I don't get this. Can't this happen with a local error process?}  The particular form of this error process is chosen to mimic a spurious error correction event, where a pair of defects is ``corrected'', but this pair of defects was not a real pair.  As time increases, the characteristic distance over which this error process could occur also increases, in accordance with the typical pair-wise separation between two defects performing a random walk.  This typical distance is exactly the factor used by the error correction algorithm to determine if a pair of defects should be corrected or not.

@@

In this section, we examine the behaviour of the protocol at different system sizes, temperatures, and measurement fractions.  System size scalings are generally reported in terms of scaling with the number of \emph{measurement patches}.  We make this choice because the size of the measurement patches is the smallest length scale relevant to the algorithm, and thus controls the majority of the scaling behavior.  Perhaps unsurprisingly, systems of radically different size but with the same number of measurement patches have fairly comparable lifetimes (e.g., size 200 system with 10 patches versus size 100 system with 10 patches).  We elaborate on the scaling of the measurement fractions further in the discussion.
\red{Define measurement fraction, $\lambda$, $\lambda_m, \lambda_b$ here.}

\begin{figure}
\begin{center}
\scalebox{1}{\includegraphics[width=1.0\columnwidth]{LifetimeVsTemperature.pdf}}
\end{center}
\caption{Lifetime of various system sizes as a function of temperature.  Notice the transition between no-effect, and significant lifetime improvement near $T=.16$. \red{I thought $\lambda$ was measurement patch size. Use a different symbol?}}
\label{fig:LifetimeVsTemperature}
\end{figure}

\figref{fig:LifetimeVsTemperature} depicts the scaling of the system lifetime with temperature.  At low temperatures, the system lifetime increases dramatically with both lower temperatures, and larger system size.  Due to finite size effects, it is difficult to extract an unambiguous threshold temperature, but below $T\approx.13$, the lifetime manifestly increases with larger system size.  \red{We empirically define a threshold temperature by ...}.

\begin{figure}
\begin{center}
\scalebox{1}{\includegraphics[width=1.0\columnwidth]{LifetimeVsSystemSize.pdf}}
\end{center}
\caption{Lifetime of the Ising model at various temperatures plotted against system size.  Note the monotonic growth in lifetime for low temperature models, as well as the plateau in lifetime for moderately sized systems.}
\label{fig:LifetimeVsSystemSize}
\end{figure}

This is clearer in \figref{fig:LifetimeVsSystemSize}, which depicts the scaling of the lifetime with system size directly.  Note that, for models above $T\approx.13$, larger system sizes asymptote to a fixed lifetime improvement, whereas for models below $T\approx.13$, the lifetime grows quasimonotonically with system size.  Beyond $\lambda\approx15$, finite size effects are significantly reduced.  This is because small-system sizes cannot easily suppress second order errors, such as defects escaping from measurement patches or multiple pairs of defects in the system.  Such errors are actually uncorrectable for systems of size $\lambda\leq4$---hence the plateau appearing around $\lambda=7$ to $\lambda=12$.  This plateau is of height $O((\Gamma_0^2)^{-1})$---the characteristic timescale of these ``second-order'' events.  This scaling is more clear when plotting the liftime divided by the bare lifetime squared---i.e. $\Gamma / \Gamma_0^2$---as in \figref{fig:LifetimeVsSystemSizeReduced}. \red{What is the measurement fraction in Figs. 2-4?}

@@

\begin{figure}
\begin{center}
\scalebox{1}{\includegraphics[width=1.0\columnwidth]{LifetimeVsMeasurementFraction.pdf}}
\end{center}
\caption{Several measurement fractions plotted as a function of temperature at fixed number of measurement patches.  Note the shift in the critical temperature as function of measurement fraction.}
\label{fig:LifetimeVsMeasurementFraction}
\end{figure}

Another free parameter in the algorithm is the ratio of measured sites $\lambda_m$ to bulk sites $\lambda_b$.  For this work, we fixed $\lambda_m=3$ and varied $\lambda_b$ \red{Move this up and define the measurement fraction in terms of these}.  \figref{fig:LifetimeVsMeasurementFraction} depicts the effect measuring fewer sites on the system lifetime as a function of temperature for several different measurement fractions.  In sum, measuring a smaller fraction of the lattice causes the threshold temperature to shift downwards.  This scaling of the threshold temperature itself is depicted explicitly in \figref{fig:CriticalTempVsMeasurementFraction}.

\begin{figure}
\begin{center}
\scalebox{1}{\includegraphics[width=1.0\columnwidth]{CriticalTempVsMeasurementFraction.pdf}}
\end{center}
\caption{The critical temperature plotted as a function of measurement fraction.  At zero measurement fraction, the critical temperature is 0, and at unit measurement fraction, the critical temperature is $\sim 1$.}
\label{fig:CriticalTempVsMeasurementFraction}
\end{figure}

\begin{figure}
\begin{center}
\scalebox{1}{\includegraphics[width=1.0\columnwidth]{LifetimeVsTemperatureVsEnergy.pdf}}
\end{center}
\caption{Here we plot the lifetime of the system as a function of temperature for a variety of energy scales $\Delta$ (see Eq. \ref{eq:isingham}).}
\label{fig:LifetimeVsTemperatureVsEnergy}
\end{figure}

@@

Finally, in Fig. \ref{fig:LifetimeVsTemperatureVsEnergy}, we plot the lifetime of the system in the presence of the protocol at various energy scales $\Delta$.  By contrasting Figures \ref{fig:LifetimeVsMeasurementFraction} and \ref{fig:LifetimeVsTemperatureVsEnergy}, one can deduce the relative benefits of error suppression via more measurement resources versus error suppression via hamiltonian engineering (i.e., a larger gap to excitation). \red{I think it would be better to put figure 7 before 6, and plot the threshold temperature as  function of m and $\Delta$ in figure 6. The this comparison can just refer to this new plot. }

@@

 In this section, we sketch how the algorithm presented in Sec. \ref{sec:ecc_alg} generalizes to a higher dimensional stabilizer quantum memory---the toric code.  Where the dynamics of the 1D Ising model are typified by one dimensional random walks of domain walls, the nonequilibrium dynamics of the toric code are driven by two dimensional random walks of ``quasiparticle'' excitations.  Consider the toric code hamiltonian:.
 \begin{align}
\Htc &= -\Delta_e \sum_v A_v -\Delta_m \sum_p B_p ,\label{eq:HTC}\\
A_v &\equiv \prod_{j \in v} \sigma_j^z,\quad B_p \equiv \prod_{j \in p} \sigma_j^x,\label{eq:AvBp}
\end{align}
\red{where $v$ denotes vertices and $p$ denotes plaquettes on a 2D square lattice} \needcite.
While domain wall excitations in the 1D ising model are associated with -1 eigenstates of the $\sigma_z^i \sigma_z^{i+1}$ stabilizers, quasiparticle excitations for the toric code are associated with -1 eigenstates of the $A_v$ and $B_p$ stabilizers as defined in Eq. \ref{eq:AvBp}.

 Broadly speaking, the algorithm is identical, but instead of having ``patches'' of measurement, there are measurement ``rails'', as indicated in \figref{fig:tc_fig}.  ``Centering'' of defects on rails amounts to shift-swaping defects into the center of the measurement rail \red{Cite your previous paper?}.  We conjecture that a sparse measurement strategy with randomly placed measurement patches of fixed diameter might exist for sufficiently large and sufficiently cold systems.  The rail geometry of \figref{fig:tc_fig} is the simplest geometry that allows us to argue for a threshold temperature for the toric code, based on a generalization of the argument used for the Ising model in Sec. \ref{sec:threshold}. \red{Reverse these two sentences and soften language in the second sentence.}

 \red{There are two types of defects in this code. Do they propagate orthogonally? If so, is each measurement rail meant to catch and center one type of defect only?}

@@

Our results complement what is known about decoders in the presence of \emph{noisy} measurements.  Figures \ref{fig:LifetimeVsMeasurementFraction} and \ref{fig:CriticalTempVsMeasurementFraction} demonstrate how a reduction in the measurement fraction in the lattice corresponds to a concomitant decrease in the threshold temperature, similar to how thresholds are known to be reduced when increasing the noise on measurements. \red{Does this suggest that one should just throw away the noisiest measurements?}

More fundamentally, our algorithm can be understood as an entropy reduction scheme.  Configurations that give rise to errors in the encoded subspace are exponentially suppressed as system size is made larger.  This is in contrast with ``energetic'' suppression---that is, suppression by widening the gap to excitations, $\Delta$ (or equivalently lowering the operating temperature).  This tradeoff between entropic and energetic contributions is depicted in Figures \ref{fig:LifetimeVsTemperatureVsEnergy} and \ref{fig:LifetimeVsMeasurementFraction}.  Depending on the resource requirements of a particular architecture, the threshold temperature can be tuned either by engineering a larger gap, $\Delta$, or by changing the number of measurements used.  In practice, this will depend on the lowest effective temperature available, the maximum measurement rate, $\chi$, \red{Is $\chi$ used anywhere else?} as well as the practical difficulty of employing more measuring devices.

@@

NEED TO DO: Argue why this equilibrium estimate is *still* a lowerbound of the nonequilibrium probability.  Intuitively it makes sense, but hard to show math-wise.

>>>>

To calculate this probability, we proceed via Bayes rule: 
\begin{align}
\label{eq:bayes}
&P(d_1:d_2 | d_1^{x_1,t_1} \wedge d_2^{x_2,t_2}) = \\ \notag
&\frac{P(d_1^{x_1,t_1} \wedge d_2^{x_2,t_2} | d_1:d_2) P(d_1 : d_2)}{P(d_1^{x_1,t_1}\wedge d_2^{x_2,t_2})}
\end{align}
The individual terms on the right hand side of equation \ref{eq:bayes} are straightforward to interpret.  $P(d_1^{x_1,t_1} \wedge d_2^{x_2,t_2} | d_1:d_2)$ represents the probability that two measured defects would be at $(x_1,t_1)$ and $(x_2,t_2)$ given that they are indeed a pair.  This can be related to the probability that a one dimensional diffusion process with diffusion constant $D$ will perform an excursion of distance $|x_2-x_1|$ or greater in a time $t_2 - t_1$; \textit{i.e.,} will perform an excursion that can measurement patches at $x_1$ and $x_2$. Explicitly,
\begin{align}
\label{eq:bayesformula}
&P(d_1^{x_1,t_1} \wedge d_2^{x_2,t_2} | d_1:d_2) \\ \notag
&= 1-2\int_{0}^{|x_2-x_1|} dx \frac{1}{2\pi D |t_2-t_1|} \rm{exp}\bigg(-\frac{|x_2-x_1|^2}{2 D |t_2-t_1|}\bigg) \\ \notag 
&= 1 - {\rm erf}\bigg(\frac{|x_2-x_1|}{2\sqrt{D |t_2 - t_1|}}\bigg) 
\end{align}
For our analysis, we will choose $D\propto \gamma_0$.  The exact correspondence between $D$ and $\gamma_0$ depends on the details of the error correction algorithm itself, so, in practice, we treat the constant of proportionality as an empirically tuned parameter.  Furthermore, we approximate any detected defects as arising from a pair that was created an equal distance between the measurement patches at locations $x_1$ and $x_2$ for the purposes of calculating the probability in Eq. \ref{eq:bayesformula}.

@@

 The simple error model is defined as follows:  suppose that a single pair of defects is in a system, and that no new pairs will be introduced.  One of the pair is fixed on a measurement rail, and the other is, at time 0, undetected.  Further, suppose that there is an error process that occasionally translates the unmeasured defect by an amount $2\sqrt{\gamma_0 \delta t}$, for hopping rate $\gamma_0$ and for system time $\delta t$.  Note that this means the error process acts on longer distance scales as system time increases.  The particular form of this error process is chosen to mimic a spurious error correction event, where a pair of defects is ``corrected'', but this pair of defects was not a real pair, leaving another unmeasured defect even further away.  As time increases, the characteristic distance over which this error process could occur also increases, in accordance with the typical pair-wise separation between two defects performing a random walk.  This typical distance is exactly the factor used by the error correction algorithm to determine if a pair of defects should be corrected or not.  Thus, the error process is representative of a worst case error correction event that not only pairs the wrong two defects, but leaves the system in a configuration with an unmeasured defect as far separated from the measured defect as possible.

@@

In this section, we examine the behaviour of the protocol at different system sizes, temperatures, and measurement fractions.  While data is reported in terms of system size, the more fundamental length scale is the characteristic size of the \emph{measurement patch}.  This is because the size of the measurement patch is the shortest length scale relevant to the algorithm, and thus controls the majority of the scaling behavior.  Perhaps unsurprisingly, systems of radically different size but with the same number of measurement patches have fairly comparable lifetimes (e.g., size 200 system with 10 patches versus size 100 system with 10 patches).  We elaborate on the scaling of the measurement fractions further in the discussion.  Unless otherwise noted, the data reported here are for systems with 3 measured sites for every 7 sites for a measurement fraction of $3/7$.  Thus, the total size of unit cell is $\lambda=7$, there are $\lambda_m=3$ measured sites, and $\lambda_b=4$ bulk sites, as is depicted in Fig. \ref{fig:detect_swap}.

\begin{figure}
\begin{center}
\scalebox{1}{\includegraphics[width=1.0\columnwidth]{LifetimeVsTemperature.pdf}}
\end{center}
\caption{Lifetime of various system sizes as a function of temperature.  Measurement fraction is $3/7$.  Notice the transition between no-effect, and significant lifetime improvement near $T=.16$.  A $3/7$ measurement fraction was used for this data.}
\label{fig:LifetimeVsTemperature}
\end{figure}

\figref{fig:LifetimeVsTemperature} depicts the scaling of the system lifetime with temperature.  At low temperatures, the system lifetime increases dramatically with both lower temperatures, and larger system size.  Due to finite size effects, it is difficult to extract an unambiguous threshold temperature, but below $T\approx.13$, the lifetime manifestly increases with larger system size.  We empirically define a threshold temperature by the choosing a large system size, fitting a curve to the Lifetime versus Temperature data which results in a lifetime improvement greater than a factor of $10$, and then back-extrapolating where this curve crosses the Temperature axis.

\begin{figure}
\begin{center}
\scalebox{1}{\includegraphics[width=1.0\columnwidth]{LifetimeVsSystemSize.pdf}}
\end{center}
\caption{Lifetime of the Ising model at various temperatures plotted against system size.  Note the monotonic growth in lifetime for low temperature models, as well as the plateau in lifetime for moderately sized systems. A $3/7$ measurement fraction was used for this data.}
\label{fig:LifetimeVsSystemSize}
\end{figure}

This is clearer in \figref{fig:LifetimeVsSystemSize}, which depicts the scaling of the lifetime with system size directly.  Note that, for models above $T\approx.13$, larger system sizes asymptote to a fixed lifetime improvement, whereas for models below $T\approx.13$, the lifetime grows quasimonotonically with system size.  Beyond $L=100 (\lambda\approx15)$, finite size effects are significantly reduced.  This is because small-system sizes cannot easily suppress second order errors, such as defects escaping from measurement patches or multiple pairs of defects in the system.  Such errors are actually uncorrectable for systems of charateristic size $\lambda\leq4$ $(L=28$ for plotted data$)$---hence the plateau appearing around $L=50 (\lambda=7)$ to $L=100 (\lambda=12)$.  This plateau is of height $O((\Gamma_0^2)^{-1})$---the characteristic timescale of these ``second-order'' events.  This scaling is more clear when plotting the liftime divided by the bare lifetime squared---i.e. $\Gamma / \Gamma_0^2$---as in \figref{fig:LifetimeVsSystemSizeReduced}.

@@

\begin{figure}
\begin{center}
\scalebox{1}{\includegraphics[width=1.0\columnwidth]{LifetimeVsMeasurementFraction.pdf}}
\end{center}
\caption{Several measurement fractions plotted as a function of temperature at fixed number of measurement patches.  Note the shift in the threshold temperature as a function of measurement fraction.}
\label{fig:LifetimeVsMeasurementFraction}
\end{figure}

Another free parameter in the algorithm is the ratio of measured sites $\lambda_m$ to bulk sites $\lambda_b$.  For this work, we fixed $\lambda_m=3$ and varied $\lambda_b$.  \figref{fig:LifetimeVsMeasurementFraction} depicts the effect measuring fewer sites on the system lifetime as a function of temperature for several different measurement fractions.  In sum, measuring a smaller fraction of the lattice causes the threshold temperature to shift downwards.  This scaling of the threshold temperature itself is depicted explicitly in \figref{fig:CriticalTempVsMeasurementFraction}.

\begin{figure}
\begin{center}
\scalebox{1}{\includegraphics[width=1.0\columnwidth]{LifetimeVsTemperatureVsEnergy.pdf}}
\end{center}
\caption{Here we plot the lifetime of the system as a function of temperature for a variety of energy scales $\Delta$ (see Eq. \ref{eq:isingham}) for an $L=224$ system.  A $3/7$ measurement fraction was used for this data.}
\label{fig:LifetimeVsTemperatureVsEnergy}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\begin{center}
\subfloat{\scalebox{1}{\includegraphics[width=\columnwidth]{CriticalTempVsMeasurementFraction.pdf}}}
\\
\vspace{-1.2\baselineskip}
\subfloat{\scalebox{1}{\includegraphics[width=\columnwidth]{CriticalTempVsDelta.pdf}}}  
\end{center}
\caption[justification=raggedright]{
Here we plot threshold temperatures as a function of measurement fraction, $m$, as well as the Hamiltonian energy scale $\Delta$ (see Eq. \ref{eq:isingham}).  At zero measurement fraction, the critical temperature is 0, and at unit measurement fraction, the critical temperature is $\sim 1$.
}
\label{fig:CriticalTempVsMeasurementFraction}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%

@@

Finally, in Fig. \ref{fig:LifetimeVsTemperatureVsEnergy}, we plot the lifetime of the system in the presence of the protocol at various energy scales $\Delta$.  By contrasting Figures \ref{fig:LifetimeVsMeasurementFraction} and \ref{fig:LifetimeVsTemperatureVsEnergy}, one can deduce the relative benefits of error suppression via more measurement resources versus error suppression via hamiltonian engineering (i.e., a larger gap to excitation).

@@

 In this section, we sketch how the algorithm presented in Sec. \ref{sec:ecc_alg} generalizes to a higher dimensional stabilizer quantum memory---the toric code.  Where the dynamics of the 1D Ising model are typified by one dimensional random walks of domain walls, the nonequilibrium dynamics of the toric code are driven by two dimensional random walks of ``quasiparticle'' excitations.  Consider the toric code hamiltonian:.
 \begin{align}
\Htc &= -\Delta_e \sum_v A_v -\Delta_m \sum_p B_p ,\label{eq:HTC}\\
A_v &\equiv \prod_{j \in v} \sigma_j^z,\quad B_p \equiv \prod_{j \in p} \sigma_j^x,\label{eq:AvBp}
\end{align}

where $v$ denotes the 4-spin vertices of the square lattice, and where and $p$ denotes the 4-qubit plaquettes on the edges of a 2D square lattice\needcite.
While domain wall excitations in the 1D ising model are associated with -1 eigenstates of the $\sigma_z^i \sigma_z^{i+1}$ stabilizers, quasiparticle excitations for the toric code are associated with -1 eigenstates of the $A_v$ and $B_p$ stabilizers as defined in Eq. \ref{eq:AvBp}.

 Broadly speaking, the algorithm is identical, but instead of having ``patches'' of measurement, there are measurement ``rails'', as indicated in \figref{fig:tc_fig}.  One such set of rails must exist for both types of excitations in the toric code---that is, one for the $B_p$ stabilizers, and another set of measurement rails for the $A_v$ stabilizers.  The error detection and correction can then be performed completely in parallel for both types of excitations, as they are independent.  ``Centering'' of defects on rails amounts to shift-swaping defects into the center of the measurement rail \red{Cite your previous paper?}.  We conjecture that a sparse measurement strategy with randomly placed measurement patches of fixed diameter might exist for sufficiently large and sufficiently cold systems.  However, the rail geometry of \figref{fig:tc_fig} is the simplest geometry that allows us to argue for a threshold temperature for the toric code, based on a generalization of the simple error model used for the Ising model in Sec. \ref{sec:threshold}.  The only difference between the upper bound to the expected scaling of the probability of uncorrectable errors in the toric code versus the expected scaling of the Ising model (i.e., Eq. \ref{eq:err_scaling}) is the prefactor becomes $\propto (L/\lambda)^2$ for the toric code instead of $L/\lambda$, where $L$ represents the linear dimension of the toric code.

@@

Our results complement what is known about decoders in the presence of \emph{noisy} measurements.  Figures \ref{fig:LifetimeVsMeasurementFraction} and \ref{fig:CriticalTempVsMeasurementFraction} demonstrate how a reduction in the measurement fraction in the lattice corresponds to a concomitant decrease in the threshold temperature, similar to how thresholds are known to be reduced when increasing the noise on measurements.

More fundamentally, our algorithm can be understood as an entropy reduction scheme.  Configurations that give rise to errors in the encoded subspace are exponentially suppressed as system size is made larger.  This is in contrast with ``energetic'' suppression---that is, suppression by widening the gap to excitations, $\Delta$ (or equivalently lowering the operating temperature).  This tradeoff between entropic and energetic contributions is depicted in Figures \ref{fig:LifetimeVsTemperatureVsEnergy} and \ref{fig:LifetimeVsMeasurementFraction}.  Depending on the resource requirements of a particular architecture, the threshold temperature can be tuned either by engineering a larger gap, $\Delta$, or by changing the number of measurements used.  In practice, this will depend on the lowest effective temperature available, the maximum measurement rate, as well as the practical difficulty of employing more measuring devices.

@@
